---
title: "Leaflet map with fire and plant intersections"
output: html_document
date: "2023-04-28"
---

Load libraries

```{r}
library(leaflet)
library(simplevis)
library(sf)
library(plyr)
library(dplyr)
library(tidyverse)
library(htmltools)
library(RColorBrewer)
```

Read in necessary files: 
1. Fire polygons 
2. Plant points (subset that overlaps with fire polygons)
3. Alaska road system
```{r}
fires <- st_read("C:\\Users\\Taylor\\Desktop\\wildfire_plotting\\AlaskaFireHistory_Polygons_1940_2021\\AlaskaFireHistory_Polygons_1940_2021\\AlaskaFireHistory_Polygons.gdb")  
fires <- subset(fires, fires$FIREYEAR > 1979)
plants <- read.csv("C:\\Users\\Taylor\\Desktop\\wildfire_plotting\\fire_plant_intersection.csv")
roads <- st_read("C:\\Users\\Taylor\\Desktop\\wildfire_plotting\\Routes.gdb\\c1808f0c-470f-48d7-b5ac-9adeef804143.gdb")
```

Now we need to clean up the plant intersection file.
```{r}
#filter out any instances of "None" or "not listed" observations
plants <- plants[!(plants$family == "None" |plants$family == "not listed"),]
#this got rid of ~3000 observations where nothing was found at a survey site (now @17k)

#filter out any observations that overlap with fires pre 1980
plants <- subset(plants, plants$FIREYEAR > 1979)
#removed about 5000 observations (now @12k)

#now we are removing duplicates (these are due to plants being recorded again if the land burned more than one time)
##order repeat values by year_diff column, should have same number of values
plants = plants[order(plants[,'sample_effort_id'],plants[,'year_diff']),] 
#filter out any duplicate instances, the most recent burn should remain
plants = plants[!duplicated(plants$sample_effort_id),]
```

Next we need to convert the `.csv` file to a `sf` object.
```{r}
plants <- st_as_sf(plants, coords = c("longitude","latitude"),crs = "NAD83")
```

Let's do some final conversions of the `sf` objects to make sure they are all in the same coordinate system. In this case we want them to be in [WGS84](https://epsg.io/4326) and the unique reference code needed is **4326**.
```{r}
roads <- st_transform(roads, 4326)
roads <- st_zm(roads) #remove z scale from the file
fires <- st_transform(fires, 4326) #only fires post 1979
plants <- st_transform(plants, 4326) #only plants overlapping with fires (this is the result of the inner join)
```

Now we are going to start working on some plotting prep!
Since we are going to color code the plant points by *rank code*, we need to get rid of any observations that don't have a ranking.

```{r}
plants <- plants[!is.na(plants$rank_code),] #started at 6557, now at 6512
```

Designate a color palette for `plants` and `fires`. (Make sure to use different ones!).We will use the library `RColorBrewer`, see color palette information [here](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/) or run `display.brewer.all()` to view options in R Studio. 
```{r}
fires_col <- colorFactor(palette = "YlOrRd",domain = fires$FIRESEASONS) #fire by fire season (decade)
plants_rank_col <- colorNumeric(palette ="GnBu", domain= fires$rank_code)#plants by rank code
```

Now we can trying creating a basic plot! *This might be a little slow.*
```{r}
ak <- leaflet() %>% addTiles() %>% 
  addPolygons(data = fires, weight = 0, fillColor = ~fires_col(FIRESEASONS), fillOpacity = 0.7) %>%
  addLegend(data = fires, pal = fires_col, values = ~FIRESEASONS,title = "Fire season", opacity = 0.7, position = "bottomright")%>%
  addCircleMarkers(data = plants, color = ~plants_rank_col(rank_code), fillOpacity = 0.7, radius = .1) %>%
  addLegend(data = plants, pal = plants_rank_col, values = ~rank_code, title = "Non-native plant </br>invasiveness ranking", position = "bottomleft")

ak
```
This plot allows you to zoom into specific regions, but you can't do much else. Let's change that.
We are going to add some popup labels to the fires and plants to make it a bit more interactive.

First, we need to create a new variable with label text in our `sf` objects.
```{r}
info_plants_labels <- paste(
  "Scientific name: ",plants$scientific_name,
  "<br>Common name: ", plants$common_name,
  "<br>Family:", plants$family,
  "<br>Invasiveness rank: ", plants$rank_code,
  "<br>Plant observed year: ", plants$year_plant,
  "<br>Fire name: ",plants$NAME_fire,
  "<br>Fire year: ", plants$FIREYEAR) %>%
  lapply(htmltools::HTML)

info_roads_labels <- paste(
roads$Route_Name_Desc_1,
"<br> Unique name:", roads$Route_Name_Unique)%>%
lapply(htmltools::HTML)

info_fires_labels <- paste(
  "Fire name:", fires$NAME,
  "<br>Year:", fires$FIREYEAR
) %>%
  lapply(htmltools::HTML)
```

Second, we need to add the labels into the map code with `popup = ~info_VARIABLE_labels`.During this step we will also add in a menu to toggle different layers on or off using `addLayersControl` and designating different layers to groups.
```{r}
ak <- leaflet() %>% addTiles(group = "Default") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Esri topo map")%>%
  addPolygons(data = fires, weight = 0, fillColor = ~fires_col(FIRESEASONS), fillOpacity = 0.7, popup = ~info_fires_labels, group = "Burned area") %>%
  addLegend(data = fires, pal = fires_col, values = ~FIRESEASONS,title = "Fire season", opacity = 0.7, position = "bottomright")%>%
  addPolylines(data = roads, color = "gray", weight = 1, popup = ~info_roads_labels, group = "Road system")%>%
  addCircleMarkers(data = plants, color = ~plants_rank_col(rank_code), fillOpacity = 0.7, radius = .1,label = ~info_plants_labels, group = "Non-native plants") %>%
  addLegend(data = plants, pal = plants_rank_col, values = ~rank_code, title = "Non-native plant </br>invasiveness ranking", position = "bottomleft") %>%
  addLayersControl(
    baseGroups = c("Default", "Esri topo map"),
    overlayGroups = c("Burned area", "Non-native plants","Road system"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% hideGroup("Road system") #hide the road system layer when this is generated
ak
```

